Improving real-time trend estimates using local parametrisation of polynomial regression filters

### 2024-08-13 paper/article_jos.qmd
doi and typos

:::: {.abstract}
This paper examines and compares real-time estimates of the trend-cycle component using moving averages constructed with local polynomial regression.
It enables to reproduce Henderson's symmetric and Musgrave's asymmetric filters used in the X-13ARIMA-SEATS seasonal adjustment algorithm.
This paper proposes two extensions of local polynomial filters for real-time trend-cycle estimates: first including a timeliness criterion to minimise the phase shift;
second with procedure for parametrising asymmetric filters locally while they are generally parametrised globally, which can be suboptimal around turning points.
An empirical comparison, based on simulated and real data, shows that modelling polynomial trends that are too complex introduces more revisions without reducing the phase shift, and that local parametrisation reduces the delay in detecting turning points and reduces revisions.
The results are reproducible and all the methods can be easily applied using the R package `rjd3filters`.

@@

Using local polynomial regression, @proietti2008 developed a general class of symmetric and asymmetric moving averages.
In particular, this class encompass Henderson's symmetric filter and Musgrave's asymmetric filters.
The aim of this study is to propose two extensions to the @proietti2008 class of filters in order to take account of two drawbacks of those methods.
Firstly  adding, in the optimisation process, a timeliness criterion to directly control the phase shift (i.e., the delay in detecting turning points).
Secondly calibrating asymmetric filters locally, while they are generally calibrated globally using thus assumptions that may be wrong locally.
Those extensions are implemented in the statistical software R [@logicielr] throw the `rjd3filters` package [@rjd3filters] and can therefore be easily used.
All the code associated to this paper is available at <https://github.com/XXXX> (anonymized linked before the manuscript is accepted).

@@

Numerous papers describe the definition and the properties of moving averages and linear filters, see for example @ch12HBSA. 
In this section we summarize some of the main results to understand the next sections.

Let $p$ and $f$ two integers, a moving average $M_{\boldsymbol\theta}$ is defined by a set of coefficients $\boldsymbol\theta=\transp{\begin{pmatrix}\theta_{-p}&\cdots&\theta_{f}\end{pmatrix}}$ such as for all time series $y_t$:
$$
M_{\boldsymbol\theta}(y_t)=\sum_{k=-p}^{+f}\theta_kX_{t+k}.
$$

@@

To sum up, applying a moving average to a harmonic times series ($y_t=\e^{-i\omega t}$) affects it in in two different ways:

@@

@@

\justifying
*Lecture: on monthly times series, 12-month cycles (associated to the frequency $2\pi/12=\pi/6$) are amplified by 10% ($G_{\boldsymbol\theta}(\pi/6)\simeq1.1$) and associated turning points are detected with a delay of 1 months ($\varphi_{\boldsymbol\theta}(\pi/6)/(\pi/6)\simeq -1$).*
*8-month cycles (associated to the irregular) are only reduced by 6% ($G_{\boldsymbol\theta}(\pi/4)\simeq0.94$).*\
*Note: the Musgrave filter depends on a ratio denoted $I/C$ (see @sec-nonparamreg).* 
*In X-11, for monthly series, when the bandwidth is 6 (i.e.: a symmetric Henderson filter of order 13 is used, what is used in most cases), the ratio is fixed to 3.5 and this value is used in this figure.*

@@

2. Apply symmetric filters to an extended forecast series (which is equivalent to using asymmetric moving averages, since forecasts are linear combinations of the past).
This method seems to date back to @deforest1877adjustment, which also suggests modelling a polynomial trend of degree three or less at the end of the period.
A similar approach is used in the X-13ARIMA-SEATS seasonal adjustment method: the series is extended by one year (by default) with an ARIMA model to minimise the revisions associated to asymmetric filters.
Since the final filter used for the seasonal adjustment extraction is much longer than one year (in most of the cases, for monthly series, the bandwidth is 78), the input series is just partially extended (and asymmetric filters are still used) to prevent from instability over long periods (e.g., outliers or model shifts).

@@

Excluding for the moment the local parametrisations of the polynomial filters, the linear-constant polynomial filter (LC) seems to give the best results in terms of delay in the detection of turning points (@fig-graphstpsimul).
Performance is relatively close to that obtained by extending the series using an ARIMA model.
However, when variability is low, the LC filter seems to give poorer results and the quadratic-linear polynomial filter (QL) seems to give the best results.

@@

\justifying
*Lecture: this plots represent box plots of the phase shift: the box represents the interquartile range, the central horizontal line the median the dot the outliers and the whisker the range of the data (if no outlier).*\
*For simulated series with medium variability (when a symmetric filter of length 13 is relevant), with the linear constant (LC) method, 50% of the turning point are detected with a phase shift of at most 5 months whereas when the LC filter is parametrised locally, 50% of the turning point are detected with a phase shift of at most 4 months.*
*For the direct asymmetric filter (DAF), 25% of the turning are detected with a phase shift of at least 7 months (compares to 5 for the LC method).*

@@

In terms of revisions, the variability of the series has little effect on the respective performances of the different methods, but it does affect the orders of magnitude, which is why the results are presented only for series with medium variability (@tbl-simulrev).
In general, LC filters always minimise revisions (with relatively small effect from the local parametrisation of the filters) and revisions are greater with cubic-quadratic (CQ) and direct (DAF) polynomial filters.

@@

\justifying
*Lecture: the estimates for $q=0$ correspond to the first estimates of the trend-cycle (real-time estimates when 0 point in the future is known), $q=1$ correspond to the second estimates of the trend-cycle (when 1 point in the future is known).*
*Since the final estimates are obtain using a symmetric filter of bandwidth 6, $q=6$ corresponds to the final estimates of the trend-cycle (when 6 points in the future are known).*
*$MAE_{fe}(0)$ compares the revision between the first estimate ($q=0$) and the last estimate ($q=6$) and $MAE_{ce}(0)$ compares the revision between the first estimate ($q=0$) and the second estimate ($q=1$).*

@@

<!-- ```{=tex} -->
<!-- \begin{table}[!h] -->
<!-- \caption{\label{tbl-simulrev} Average of the relative deviations of the revisions for the different -->
<!-- filters on simulated series with medium variability.} -->
<!-- \centering -->
<!-- \begin{tabular}{ccccccc} -->
<!-- \toprule -->
<!-- Method & $q=0$ & $q=1$ & $q=2$ & $q=3$ & $q=4$ & $q=5$\\ -->
<!-- \midrule -->
<!-- \addlinespace[0.3em] -->
<!-- \multicolumn{7}{l}{\textbf{$MAE_{fe}(q) = \mathbb E\left[\left|(TC_{t|t+q} -  TC_{t|last})/TC_{t|last}\right|\right]$}}\\ -->
<!-- \hspace{1em}LC & 0.21 & 0.10 & 0.03 & 0.03 & 0.03 & 0.01\\ -->
<!-- \hspace{1em}LC local param. (final estimates) & 0.19 & 0.09 & 0.03 & 0.03 & 0.03 & 0.01\\ -->
<!-- \hspace{1em}LC local param. & 0.29 & 0.10 & 0.03 & 0.03 & 0.03 & 0.01\\ -->
<!-- \hspace{1em}QL & 0.33 & 0.10 & 0.04 & 0.04 & 0.03 & 0.01\\ -->
<!-- \hspace{1em}QL local param. (final estimates) & 0.21 & 0.10 & 0.03 & 0.03 & 0.03 & 0.01\\ -->
<!-- \hspace{1em}QL local param. & 0.30 & 0.10 & 0.04 & 0.03 & 0.03 & 0.01\\ -->
<!-- \hspace{1em}CQ & 0.45 & 0.13 & 0.13 & 0.09 & 0.06 & 0.02\\ -->
<!-- \hspace{1em}DAF & 0.47 & 0.15 & 0.15 & 0.09 & 0.06 & 0.02\\ -->
<!-- \hspace{1em}ARIMA & 0.22 & 0.10 & 0.03 & 0.03 & 0.03 & 0.01\\ -->
<!-- \addlinespace[0.3em] -->
<!-- \multicolumn{7}{l}{\textbf{$MAE_{ce}(q)=\mathbb E\left[ -->
<!-- \left|(TC_{t|t+q} - TC_{t|t+q+1})/TC_{t|t+q+1}\right| -->
<!-- \right]$}}\\ -->
<!-- \hspace{1em}LC & 0.19 & 0.10 & 0.02 & 0.01 & 0.07 & 0.01\\ -->
<!-- \hspace{1em}LC local param. (final estimates) & 0.20 & 0.10 & 0.03 & 0.01 & 0.05 & 0.01\\ -->
<!-- \hspace{1em}LC local param. & 0.24 & 0.11 & 0.03 & 0.01 & 0.05 & 0.01\\ -->
<!-- \hspace{1em}QL & 0.29 & 3.46 & 0.00 & 0.03 & 0.04 & 0.01\\ -->
<!-- \hspace{1em}QL local param. (final estimates) & 0.31 & 0.11 & 0.02 & 0.01 & 0.04 & 0.01\\ -->
<!-- \hspace{1em}QL local param. & 0.24 & 0.16 & 0.00 & 0.03 & 0.04 & 0.01\\ -->
<!-- \hspace{1em}CQ & 0.43 & 0.02 & 0.10 & 0.07 & 0.05 & 0.02\\ -->
<!-- \hspace{1em}DAF & 0.66 & 0.24 & 0.11 & 0.14 & 0.06 & 0.02\\ -->
<!-- \hspace{1em}ARIMA & 0.21 & 0.13 & 0.02 & 0.02 & 0.25 & 0.01\\ -->
<!-- \bottomrule -->
<!-- \end{tabular} -->
<!-- \end{table} -->
<!-- ``` -->

@@

\justifying
*Lecture: the dotted vertical line corresponds to the date of the turning point (February 2001).*
*The curve "Feb 2001" corresponds to the estimates of the trend-cycle using the data observed until February 2001.*
*Since a 13-term Henderson filter is used for the final estimates, for the curve "Feb 2001" estimates from September 2001 to February 2001 are intermediate estimates.*

Successive estimates of the trend-cycle in US employment (in logarithms).
:::

@@

\justifying
*Lecture: the dotted vertical line corresponds to the date of the turning point (February 2001).*
*The curve "Mar 2001" corresponds to the implicit forecasts associated to the asymmetric filters used for the trend-cycle estimations using the data observed until March 2001.*

Implicit forecasts linked to successive estimates of the trend-cycle in US employment (in logarithms) using local polynomial methods.
:::

@@

Since X-13ARIMA-SEATS can only be applied for series with frequency lower than monthly, for higher frequencies (weekly, daily, etc.), there is no reference on the length of the filters to uses.
The length of the filters can be determined using local polynomial methods, for example minimising a cross-validation statistics (using the function `rjd3filters::cv()`) or with more complex methods as describe by @Loader1999, but it will surely be necessary to use higher bandwidth than for monthly series.
The larger the window, the greater the bias in assuming that the trend is locally of degree 1: the LC method will then be certainly less appropriate than other methods (QL, CQ or DAF).
Moreover, since the bias is higher, the local parametrisation of the filters will surely have more impact.

@@

@Fig-ce16ovcovid and @fig-ce16ov-previmp-covid show the successive estimates of the trend-cycle and the associated implicit forecast for the US employment during the COVID-19.
First, we observe that while the real turning point is in April 2020, the 13-term Henderson filter produces final estimates of the trend-cycles, biased by the presence of a huge outlier, which lead to a turning point detected in June 2020.
Second, as in 2001, the CQ and DAF methods lead to lots of variability in the intermediate estimates.
For the QL method, the local parametrisation seems to slightly reduce the revisions.
For the LC method, local parametrisation has few impact on the estimates of the trend-cycle but the associated implicit forecasts are not plausible: the final estimate of the slope is biased by the outlier.
Third, the ARIMA model can lead to unrealistic forecasts.

@@

- Since all the filter studied in this paper are equivalent to a local polynomial model, an external regressor can be added to the linear model to handle the outlier (the matrix $\boldsymbol X$ would then be modified).
In the case of an additive outlier, this is equivalent to set to 0 the coefficient of the moving average associated to the date of the outlier (using $y_{t-h}$ to $y_{t+h}$ for the trend-cycle estimates, if the outlier is at the date $t$ the central coefficient is set to 0) and renormalise the sum of the coefficients to 1.

>>>>

:::: {.abstract}
This paper examines and compares real-time estimates of the trend-cycle component using moving averages constructed with local polynomial regression.
It enables the reproduction of Henderson's symmetric and Musgrave's asymmetric filters used in the X-13ARIMA-SEATS seasonal adjustment algorithm.
This paper proposes two extensions of local polynomial filters for real-time trend-cycle estimates: first including a timeliness criterion to minimise the phase shift;
second with procedure for parametrising asymmetric filters locally while they are generally parametrised globally, which can be suboptimal around turning points.
An empirical comparison, based on simulated and real data, shows that modelling polynomial trends that are too complex introduces more revisions without reducing the phase shift, and that local parametrisation reduces the delay in detecting turning points and reduces revisions.
The results are reproducible and all the methods can be easily applied using the R package `rjd3filters`.

@@

Using local polynomial regression, @proietti2008 developed a general class of symmetric and asymmetric moving averages.
In particular, this class encompass Henderson's symmetric filter and Musgrave's asymmetric filters.
The aim of this study is to propose two extensions to the @proietti2008 class of filters in order to take account of two drawbacks of those methods: 
firstly  adding, in the optimisation process, a timeliness criterion to directly control the phase shift (i.e., the delay in detecting turning points);
secondly calibrating asymmetric filters locally, while they are generally calibrated globally using assumptions that may be wrong locally.
Those extensions are implemented in the statistical software R [@logicielr] through the `rjd3filters` package [@rjd3filters] and can therefore be easily used.
All the code associated to this paper is available at <https://github.com/AQLT/joffstats2024>.

@@

Numerous papers describe the definition and the properties of moving averages and linear filters; see for example @ch12HBSA. 
In this section we summarize some of the main results to understand the next sections.

Let $p$ and $f$ two integers, a moving average $M_{\boldsymbol\theta}$ is defined by a set of coefficients $\boldsymbol\theta=\transp{\begin{pmatrix}\theta_{-p}&\cdots&\theta_{f}\end{pmatrix}}$ such that for all time series $y_t$:
$$
M_{\boldsymbol\theta}(y_t)=\sum_{k=-p}^{+f}\theta_kX_{t+k}.
$$

@@

To sum up, applying a moving average to a harmonic times series ($y_t=\e^{-i\omega t}$) affects it in two different ways:

@@

Observed in @fig-graphsmusgrave that for monthly times series, 12-month cycles (associated to the frequency $2\pi/12=\pi/6$) are amplified by 10% ($G_{\boldsymbol\theta}(\pi/6)\simeq1.1$) and associated turning points are detected with a delay of 1 months ($\varphi_{\boldsymbol\theta}(\pi/6)/(\pi/6)\simeq -1$).
8-month cycles (associated to the irregular) are only reduced by 6% ($G_{\boldsymbol\theta}(\pi/4)\simeq0.94$).
Note that the Musgrave filter depends on a ratio denoted $I/C$ (see @sec-nonparamreg).
In X-11, for monthly series, when the bandwidth is 6 (i.e.: a symmetric Henderson filter of order 13 is used, what is used in most cases), the ratio is fixed to 3.5 and this value is used in this figure.

@@

@@

2. Apply symmetric filters to an extended forecast series (which is equivalent to using asymmetric moving averages, since forecasts are linear combinations of the past).
This method seems to date back to @deforest1877adjustment, which also suggests modelling a polynomial trend of degree three or less at the end of the period.
A similar approach is used in the X-13ARIMA-SEATS seasonal adjustment method: the series is extended by one year (by default) with an ARIMA model to minimise the revisions associated to asymmetric filters.
Since the final filter used for the seasonal adjustment extraction is much longer than one year (in most of the cases, for monthly series, the bandwidth is 78), the input series is just partially extended (and asymmetric filters are still used) to prevent instability over long periods (e.g., outliers or model shifts).

@@

@Fig-graphstpsimul represents box plots of the phase shift: the box represents the interquartile range, the central horizontal line the median the dot the outliers and the whisker the range of the data (if no outlier).
For simulated series with medium variability (when a symmetric filter of length 13 is relevant), with the linear constant (LC) method, 50% of the turning point are detected with a phase shift of at most 5 months whereas when the LC filter is parametrised locally, 50% of the turning point are detected with a phase shift of at most 4 months.
For the direct asymmetric filter (DAF), 25% of the turning are detected with a phase shift of at least 7 months (compares to 5 for the LC method).

Excluding for the moment the local parametrisations of the polynomial filters, the linear-constant polynomial filter (LC) seems to give the best results in terms of delay in the detection of turning points.
Performance is relatively close to that obtained by extending the series using an ARIMA model.
However, when variability is low, the LC filter seems to give poorer results and the quadratic-linear polynomial filter (QL) seems to give the best results.

@@

@@

Observe @tbl-simulrev the average of the relative deviations with medium variability. 
The estimates for $q=0$ correspond to the first estimates of the trend-cycle (real-time estimates when 0 point in the future is known), $q=1$ correspond to the second estimates of the trend-cycle (when 1 point in the future is known).
Since the final estimates are obtain using a symmetric filter of bandwidth 6, $q=6$ corresponds to the final estimates of the trend-cycle (when 6 points in the future are known).
$MAE_{fe}(0)$ compares the revision between the first estimate ($q=0$) and the last estimate ($q=6$) and $MAE_{ce}(0)$ compares the revision between the first estimate ($q=0$) and the second estimate ($q=1$).

In terms of revisions, the variability of the series has little effect on the respective performances of the different methods, but it does affect the orders of magnitude, which is why the results are presented only for series with medium variability.
In general, LC filters always minimise revisions (with relatively small effect from the local parametrisation of the filters) and revisions are greater with cubic-quadratic (CQ) and direct (DAF) polynomial filters.

@@

@@

@@

Successive estimates of the trend-cycle in US employment (in logarithms). 
The dotted vertical line corresponds to the date of the turning point (February 2001).
The curve "Feb 2001" corresponds to the estimates of the trend-cycle using the data observed until February 2001.
Since a 13-term Henderson filter is used for the final estimates, for the curve "Feb 2001" estimates from September 2001 to February 2001 are intermediate estimates.
:::

@@

Implicit forecasts linked to successive estimates of the trend-cycle in US employment (in logarithms) using local polynomial methods.
The dotted vertical line corresponds to the date of the turning point (February 2001).
The curve "Mar 2001" corresponds to the implicit forecasts associated to the asymmetric filters used for the trend-cycle estimations using the data observed until March 2001.
:::

@@

Since X-13ARIMA-SEATS can only be applied for series with frequency lower than monthly, for higher frequencies (weekly, daily, etc.), there is no reference on the length of the filters to uses.
The length of the filters can be determined using local polynomial methods, for example minimising a statistic such as cross-validation (`rjd3filters::cv()`), CP (`rjd3filters::cp()`) or Rice's T (`rjd3filters::rt()`), or with more complex methods as describe by @Loader1999, but it will surely be necessary to use higher bandwidth than for monthly series.
The larger the window, the greater the bias in assuming that the trend is locally of degree 1: the LC method will then be certainly less appropriate than other methods (QL, CQ or DAF).
Moreover, since the bias is higher, the local parametrisation of the filters will surely have more impact.

@@

@Fig-ce16ovcovid and @fig-ce16ov-previmp-covid show the successive estimates of the trend-cycle and the associated implicit forecast for the US employment during the COVID-19.
First, we observe that while the real turning point is in April 2020, the 13-term Henderson filter produces final estimates of the trend-cycles, biased by the presence of a huge outlier, which lead to a turning point detected in June 2020.
Second, as in 2001, the CQ and DAF methods lead to lots of variability in the intermediate estimates.
For the QL method, the local parametrisation seems to slightly reduce the revisions.
For the LC method, local parametrisation has little impact on the estimates of the trend-cycle but the associated implicit forecasts are not plausible: the final estimate of the slope is biased by the outlier.
Third, the ARIMA model can lead to unrealistic forecasts.

@@

- Since all the filters studied in this paper are equivalent to a local polynomial model, an external regressor can be added to the linear model to handle the outlier (the matrix $\boldsymbol X$ would then be modified).
In the case of an additive outlier, this is equivalent to setting to 0 the coefficient of the moving average associated to the date of the outlier (using $y_{t-h}$ to $y_{t+h}$ for the trend-cycle estimates, if the outlier is at the date $t$ the central coefficient is set to 0) and renormalise the sum of the coefficients to 1.
